Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-23 Manuscript/sec4-experiments.tex
update by www.grammarly.com

We answer the four questions in Section~\ref{sec:experimental_scheme} one by one with experiments.
Without otherwise mentioned, the reported training time per epoch is the average wall-clock training time of 50 epoches, excluding abnormal epoches \footnote{During the training of some epoches, there were extra profiling overheads from NVIDIA Nsight Systems and GC pauses from the Python interpreter that significantly increased the training time. Assume Q1 and Q3 are the 25\% and 75\% quantiles of the training time of 50 epoches, respectively. We regard the epoches with the training time \emph{outside} the range of $[Q1 - 1.5 * (Q3-Q1), Q3 + 1.5 * (Q3-Q1)]$ as abnormal epoches.}.

@@

For GaAN, it is also based on the multi-head mechanism.
Its time complexity is affected by $d_{in}$, $d_v$, $d_a$ and the number of heads $K$.
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gaan} demonstrates that the training time increased linearly with the hyper-parameters, except for $dim(\boldsymbol{h}^1)$.
As $dim(\boldsymbol{h}^1)$ increased, the training time increased first slightly and then linearly.
We could observe similar phenomena in GCN, GGNN and GAT with low hyper-parameters.
When the hyper-parameter was too low, the GNN training could not make use of the full computing power of the GPU.
When it became high enough, the training time increased linearly.

@@

The experimental results support the time complexity analysis.
We further measured the effects of the hyper-parameters on the peak GPU memory usage in \figurename~\ref{fig:exp_hyperparameter_memory_usage}.
The memory usage also increased linearly as the hyper-parameters increased for all GNNs, except for GaAN on $dim(\boldsymbol{h}^1)$.
As the hidden feature vectors $\boldsymbol{h}^1$ consumed a small proportion of memory in GaAN, the growth in the memory usage was not noticable until $dim(\boldsymbol{h}^1)$ was large enough.

@@

The complexity analysis in \tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} is valid.
The hyper-parameters affect the training time and the memory usage of the GNN training \textbf{in a linear way}.
Algorithm engineers can use larger hyper-parameters to increase the expressive power of a GNN model without worrying the explosive growth in the training time and memory usage.

@@

\figurename~\ref{fig:exp_vertex_edge_cal_proportion} decomposes the training time of a GNN on the layer level.
The training time of each layer is the summation of the time in the forward, backward and evaluation phases.
In GCN, GAT and GaAN, the time spent on the layer 0 was much larger than the layer 1.
In those GNNs, the dimensions of the input/output feature vectors in the layer 0 were much larger than the dimensions in the layer 1.
$d^0_{in}=dim(\boldsymbol{x})$, $d^0_{out}=d^1_{in}=64$ and $d^1_{out}=\#Class$ and $dim(\boldsymbol{x}) \gg \#Class$.
For GaAN, since it required the dimensions of the input/output feature vectors must be same, $d^0_{in}=d^0_{out}=d^1_{in}=d^1_{out}=64$ and the training time of both layers were close.

Each GNN layer can be further divided in the vertex and the edge calculation stages.
In \figurename~\ref{fig:exp_vertex_edge_cal_proportion}, GCN spent most of the training time on the edge calculation in most datasets.
A special case is \texttt{cph} dataset.
The dimension of the input feature vectors was very high in \texttt{cph}, making the vertex calculation stage of the GCN Layer 0 spend considerable time.
GGNN also spent the majority of its training time on the edge calculation.
But the high time complexity of its vertex update function $\gamma$ made the ratio of the vertex calculation in the total training time much higher than other GNNs.
%In \textit{pub} and cam dataset, the edge calculation cost and the vertex calculation cost are close for GGNN
%because the average degree of the two datasets is low (only 4.5 and 2.8).
For GAT and GaAN, due to their high edge calculation complexity, the edge calculation is the absolutely dominant stage.
In summary, \emph{the edge calculation is the most time-consuming stage in the GNN training}.

@@

The experimental results also indicate that the average degree of the dataset affects the time-consuming proportion of the vertex/edge calculation.
For GaAN, the time spent on the vertex calculation exceeded the edge calculation on the \texttt{pub} and \texttt{cam} datasets, because the avaerage degrees of the two datasets were low, making $|\mathcal{E}|$ and $|\mathcal{V}|$ much closer.
To evaluate the effects of the average degree, we used the R-MAT model to generate random graphs with 50k vertices and the average degrees ranging from 2 to 100.
\figurename~\ref{fig:exp_avg_degree_on_vertex_edge_cal_time} shows the training time of the four GNNs under different average degrees.
As the average degree increased, the training time of the edge calculation grew \emph{linearly}.
For GCN, GAT and GaAN, the edge calculation dominated the entire training time even under small average degrees.
Only for GGNN that had high vertex and low edge calculation complexities, the training time of the vertex calculation could exceed the edge calculation under low average degrees ($<5$).
Therefore, \emph{improving the efficiency of the edge calculation is the key to reduce the GNN training time}.

@@

In the implementation of PyG, the edge calculation stage can be decomposed into four steps: collect, message, aggregate and update, as shown in \figurename~\ref{fig:steps_in_edge_calculation}.
The edge index is a matrix with $M$ rows and 2 columns that holds the edge set of the graph, where $M=|\mathcal{E}|$.
The two columns of the matrix store the source vertex and the target vertex of each edge, respectively.
The collect step copies the vertex feature vectors from the previous layer $\boldsymbol{h}_i^l$ to the ends of each edge
in the edge index, to form the parameters $[\boldsymbol{h}^l_i, \boldsymbol{h}^l_{j}, \boldsymbol{e}^l_{i,j}]$ of the message function $\phi$.
This step only involves the data movement.
The message step calls the message function $\phi$ to get message vectors of all edges $\boldsymbol{m}_{i, j}^l$.
The aggregate step aggregates the message vectors with the same target vertex into a aggregated vector $\boldsymbol{s}^l_i$ with the aggregation operation $\Sigma$.
The update step is optinal.
It performs additional transformation on the aggregated vectors (for example, adding bias in GCN).
The aggregated vectors $\boldsymbol{s}^l_i$ (after the update step) will be fed into the vertex update function $\gamma$ as one of the input parameters.

@@

We decomposed the execution time of the edge calculation step in \figurename~\ref{fig:exp_edge_calc_decomposition}.
In each GNN, the proportion of the four steps were rather stable, rarely affected by datasets.
For GAT and GaAN with the high edge calculation complexity, the message step consumed most of the execution time.
For GCN and GGNN with the low complexity, the proportions of the steps were close.
Since the message function $\phi$ of GGNN used the pre-computed $\hat{\boldsymbol{h}}^l_i$ as the message vector directly in the implementation, the time spent on the message step of GGNN was negligible.
Although the collect step did not conduct any computation and only involved data movement, it occupied noticeable execution time in all the GNNs.
The experiments show that \emph{the performance bottleneck on the step level depends on the complexity of the edge calculation}.
For GNNs with the high edge calculation complexity, the message function $\phi$ is the performance bottleneck.
Optimizing its implementation can significantly reduce the training time.
For the other GNNs, optimization should focus on reducing the costs of the collect and the aggregation steps.
Addtionally, improving the efficiency of the collect step can benefit all GNNs.

@@

\paragraph{GGNN}
The top basic operator was \texttt{mm} used in the vertex calculation.
Due to the high time complexity in $\gamma$, the proportion of the \texttt{mm} operator were much higher than the other operators.
The \texttt{thnn\_fused\_gru\_cell} operator that was used in the backward phase of $\gamma$ was also noticeable.
The other three operators were used in the edge calculation.

@@

\paragraph{Summary}
The GNN training is suitable for GPU.
The \textbf{edge calculation is the main performance bottleneck}, except for training GNNs with high vertex calculation complexity on low-average-degree graphs.
The performance bottleneck in the edge calculation depends on the time complexity of the message function $\phi$.
\begin{itemize}
    \item If the time complexity of $\phi$ is \textbf{high}, the \textbf{efficiency of $\phi$} limits the performance. Reducing its computation cost (via optimizing its implementation or modifying the algorithm) can significantly reduce the training time.
    \item If the time complexity of $\phi$ is \textbf{low}, the \textbf{collect step} and the \textbf{aggregation step} limit the performance. The collect step involves lots of data movement. The aggregation step suffers from the data synchronization and non-regular data access. Optimizing their implementations can significantly reduce the training time.
\end{itemize}

@@

\figurename~\ref{fig:exp_memory_usage_stage_amp} shows the peak memory usage of each phase in the GNN training on the \texttt{amp} dataset.
The trend was similar on other datasets.
\emph{The GNN training achieved its peak memory usage in the forward and the backward phases}.
The forward phase generated a large number of intermediate results.
Some key intermediate results were cached, increasing the memory usage.
The cached results were used in the gradient calculation in the backward phase.
\figurename~\ref{fig:ggnn_vertex_func_computation_graph} shows the computation graph of the vertex update function $\gamma$ of GGNN.
The computation graph had a large number of operators, each operator generating an intermediate matrix.
Some key intermediate matrices were cached.
The cached matrices were the main source of memory usage in the loss phase.
By the end of the backward phase, the cached results were released.
Since the evaluation phase did not need to calculate the gradients, it did not cache intermediate results.
Its memory usage declined sharply.

@@

\figurename~\ref{fig:exp_memory_expansion_ratio} also indicates that the same GNN had different MERs for different datasets.
Two characteristics of the dataset affects the MER: the dimension of the input feature vector and the average degree.

@@

The average degree of the graph also affects MER by influencing the relative scale of the intermediate results from vertex/edge calculation.
Fixing the number of vertices $|\mathcal{V}|$, we used the R-MAT generator to generate random graphs with different average degrees.
\figurename~\ref{fig:exp_memory_expansion_ratio_input_graph_number_of_edges} shows how the memory usage changes according to the average degrees.
As the average degree $\bar{d}$ increased, the number of edges $|\mathcal{E}|$ increased and the peak memory usage increased \emph{linearly} with $\bar{d}$.
The edge calculation gradually dominated the memory usage and \emph{the MER converged to a stable value}.
The stable value was determined by the complexity of the edge calculation.
Except for GGNN, the MERs of the other GNNs increased as $\bar{d}$ increased.
As GGNN had high vertex calculation complexity, the MER related to the vertex calculation was much larger than the edge calculation.
When the edge calculation dominated the memory usage, its MER became smaller.

@@

We also fixed the number of edges $|\mathcal{E}|$ in the graph and generated random graphs with different $|\mathcal{V}|$.
\figurename~\ref{fig:exp_memory_expansion_ratio_input_graph_number_of_vertices_fixed_edge} shows how the memory usage changed according to $|\mathcal{V}|$.
All GNNs were much more insensitive to the changes in $|\mathcal{V}|$ compared to $|\mathcal{E}|$.
Except for GGNN, the MERs of the other GNNs declined as $|\mathcal{V}|$ increased because the scale of the dataset increased quicklier than the scales of the intermediate results.
As GGNN had high vertex calculation complexity, the scales of the intermediate results were much more sensitive to $|\mathcal{V}|$.
It indicates that \emph{the intermediate results of the edge calculation dominated the memory usage during the GNN training}.

@@

In the implementation in PyG, the GNN model and the dataset resided on the GPU side.
To process each epoch, PyG sampled the original dataset in the main memory and generated several batches.
Each batch was a small subgraph of the dataset.
To train on each batch, PyG sent the sampled subgraph to GPU, calculated the gradients in the subgraph and updated the model parameters directly on GPU.
With the sampling techniques, the model parameters were updated by a stochastic gradient descent optimizer.
It conducted the evaluation phase every several epoches (either on the CPU side or the GPU side) to determine whether to stop the training.
The experiments focused on the training phase of each batch.

@@

\figurename~\ref{fig:exp_sampling_minibatch_graph_info} shows how the size of the sampled subgraph changed with the batch size.
For the neighbor sampler, the relative batch size is the porportion of the sampled vertices of the last GNN layer in all vertices of the graph.
For the cluster sampler, the relative batch size is the proportion of the sampled partitions to all partitions of the graph.
The neighbor sampler was very sensitive to the batch size.
As the batch size increased, the size of the sampled subgraph increased quickly and then stabalized.
The cluster sampler was much less sensitive compared to the neighbor sampler.
The number of vertices and the average degree of the sampled subgraphs increased linearly with the batch size.

@@

To find out the performance bottleneck with the sampling techniques, we decomposed the training time per batch into three phases: \emph{sampling} on CPU, \emph{transfering} sampled subgraphs from CPU to GPU and \emph{training} with the subgraphs on GPU.
\figurename~\ref{fig:exp_sampling_batch_train_time} shows the time breakdown of the four GNNs under different relative batch sizes.
For the neighbor sampler, the sampling technique reduced the training time per batch only when the batch size was very small.
When the batch became bigger, the sampling and the data transfering phases introduced noticeable overheads, even making the training time exceed the full-batch training.
For the clustering sampler, the sampled subgraph was smaller than the neighbor sampler under the same relative batch size.
The reduction in the training time was more obvious than the neighbor sampler.
However, the overheads increased quickly as the relative batch size increased.
The training time under the 25\% relative batch size already exceeded the time of full-batch training.
The experimental results indicate that the implementation of the sampling techniques in PyG is inefficient.
When the batch size was slightly big, more than 50\% of the time had been spent on sampling and data transfering.
\emph{The sampling techniques are only efficient under small batch sizes.}

@@

The disadvantage of the sampling technique is wasting the GPU resources.
As the sampling techniques are only effective under small batch sizes, the sampled subgraphs will be very small in those cases.
They cannot make full use of the computing power of a GPU.
To simulate the cases of training GNNs with small sampled subgraphs, we generated random graphs with few vertices and measured the training time per epoch in \figurename~\ref{fig:exp_small_graph_train_time}.
As the number of vertices increased, the training time was almost unchanged except for GaAN.
The training time of GaAN increased only with $|\mathcal{V}| \geq 4000$.

\paragraph{Summary}
With small batch size, the sampling techniques could significantly reduce the training time per batch and the peak memory usage during the training.
However, small batch sizes could not make full use of the computing power of a GPU.
The sampled subgraphs had lower average degrees than the whole graph.
However, the current implementation of the sampling techniques was every inefficient under big batch sizes.
The time spent on the sampling phase and the data transfering phase could even exceeded the training phase.

>>>>

We answer the four questions in Section~\ref{sec:experimental_scheme} one by one with experiments.
Without otherwise mentioned, the reported training time per epoch is the average wall-clock training time of 50 epochs, excluding abnormal epochs \footnote{During the training of some epochs, there were extra profiling overheads from NVIDIA Nsight Systems and GC pauses from the Python interpreter that significantly increased the training time. Assume Q1 and Q3 are the 25\% and 75\% quantiles of the training time of 50 epochs, respectively. We regard the epochs with the training time \emph{outside} the range of $[Q1 - 1.5 * (Q3-Q1), Q3 + 1.5 * (Q3-Q1)]$ as abnormal epochs.}.

@@

For GaAN, it is also based on the multi-head mechanism.
Its time complexity is affected by $d_{in}$, $d_v$, $d_a$ and the number of heads $K$.
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gaan} demonstrates that the training time increased linearly with the hyper-parameters, except for $dim(\boldsymbol{h}^1)$.
As $dim(\boldsymbol{h}^1)$ increased, the training time increased first slightly and then linearly.
We could observe similar phenomena in GCN, GGNN, and GAT with low hyper-parameters.
When the hyper-parameter was too low, the GNN training could not make use of the full computing power of the GPU.
When it became high enough, the training time increased linearly.

@@

The experimental results support the time complexity analysis.
We further measured the effects of the hyper-parameters on the peak GPU memory usage in \figurename~\ref{fig:exp_hyperparameter_memory_usage}.
The memory usage also increased linearly as the hyper-parameters increased for all GNNs, except for GaAN on $dim(\boldsymbol{h}^1)$.
As the hidden feature vectors $\boldsymbol{h}^1$ consumed a small proportion of memory in GaAN, the growth in the memory usage was not noticeable until $dim(\boldsymbol{h}^1)$ was large enough.

@@

The complexity analysis in \tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} is valid.
The hyper-parameters affect the training time and the memory usage of the GNN training \textbf{in a linear way}.
Algorithm engineers can use larger hyper-parameters to increase the expressive power of a GNN model without worrying about the explosive growth in the training time and memory usage.

@@

\figurename~\ref{fig:exp_vertex_edge_cal_proportion} decomposes the training time of a GNN on the layer level.
The training time of each layer is the summation of the time in the forward, backward, and evaluation phases.
In GCN, GAT, and GaAN, the time spent on layer 0 was much larger than layer 1.
In those GNNs, the dimensions of the input/output feature vectors in the layer 0 were much larger than the dimensions in the layer 1.
$d^0_{in}=dim(\boldsymbol{x})$, $d^0_{out}=d^1_{in}=64$ and $d^1_{out}=\#Class$ and $dim(\boldsymbol{x}) \gg \#Class$.
For GaAN, since it required the dimensions of the input/output feature vectors must be the same, $d^0_{in}=d^0_{out}=d^1_{in}=d^1_{out}=64$ and the training time of both layers were close.

Each GNN layer can be further divided into the vertex and the edge calculation stages.
In \figurename~\ref{fig:exp_vertex_edge_cal_proportion}, GCN spent most of the training time on the edge calculation in most datasets.
A special case is \texttt{cph} dataset.
The dimension of the input feature vectors was very high in \texttt{cph}, making the vertex calculation stage of the GCN Layer 0 spend considerable time.
GGNN also spent the majority of its training time on the edge calculation.
But the high time complexity of its vertex update function $\gamma$ made the ratio of the vertex calculation in the total training time much higher than other GNNs.
%In \textit{pub} and cam dataset, the edge calculation cost, and the vertex calculation cost are close for GGNN
%because the average degree of the two datasets is low (only 4.5 and 2.8).
For GAT and GaAN, due to their high edge calculation complexity, the edge calculation is the absolutely dominant stage.
In summary, \emph{the edge calculation is the most time-consuming stage in the GNN training}.

@@

The experimental results also indicate that the average degree of the dataset affects the time-consuming proportion of the vertex/edge calculation.
For GaAN, the time spent on the vertex calculation exceeded the edge calculation on the \texttt{pub} and \texttt{cam} datasets, because the average degrees of the two datasets were low, making $|\mathcal{E}|$ and $|\mathcal{V}|$ much closer.
To evaluate the effects of the average degree, we used the R-MAT model to generate random graphs with 50k vertices and the average degrees ranging from 2 to 100.
\figurename~\ref{fig:exp_avg_degree_on_vertex_edge_cal_time} shows the training time of the four GNNs under different average degrees.
As the average degree increased, the training time of the edge calculation grew \emph{linearly}.
For GCN, GAT, and GaAN, the edge calculation dominated the entire training time even under small average degrees.
Only for GGNN that had high vertex and low edge calculation complexities, the training time of the vertex calculation could exceed the edge calculation under low average degrees ($<5$).
Therefore, \emph{improving the efficiency of the edge calculation is the key to reduce the GNN training time}.

@@

In the implementation of PyG, the edge calculation stage can be decomposed into four steps: collect, message, aggregate, and update, as shown in \figurename~\ref{fig:steps_in_edge_calculation}.
The edge index is a matrix with $M$ rows and 2 columns that holds the edge set of the graph, where $M=|\mathcal{E}|$.
The two columns of the matrix store the source vertex and the target vertex of each edge, respectively.
The collect step copies the vertex feature vectors from the previous layer $\boldsymbol{h}_i^l$ to the ends of each edge
in the edge index, to form the parameters $[\boldsymbol{h}^l_i, \boldsymbol{h}^l_{j}, \boldsymbol{e}^l_{i,j}]$ of the message function $\phi$.
This step only involves the data movement.
The message step calls the message function $\phi$ to get message vectors of all edges $\boldsymbol{m}_{i, j}^l$.
The aggregate step aggregates the message vectors with the same target vertex into an aggregated vector $\boldsymbol{s}^l_i$ with the aggregation operation $\Sigma$.
The update step is optional.
It performs an additional transformation on the aggregated vectors (for example, adding bias in GCN).
The aggregated vectors $\boldsymbol{s}^l_i$ (after the update step) will be fed into the vertex update function $\gamma$ as one of the input parameters.

@@

We decomposed the execution time of the edge calculation step in \figurename~\ref{fig:exp_edge_calc_decomposition}.
In each GNN, the proportion of the four steps were rather stable, rarely affected by datasets.
For GAT and GaAN with the high edge calculation complexity, the message step consumed most of the execution time.
For GCN and GGNN with the low complexity, the proportions of the steps were close.
Since the message function $\phi$ of GGNN used the pre-computed $\hat{\boldsymbol{h}}^l_i$ as the message vector directly in the implementation, the time spent on the message step of GGNN was negligible.
Although the collect step did not conduct any computation and only involved data movement, it occupied noticeable execution time in all the GNNs.
The experiments show that \emph{the performance bottleneck on the step level depends on the complexity of the edge calculation}.
For GNNs with the high edge calculation complexity, the message function $\phi$ is the performance bottleneck.
Optimizing its implementation can significantly reduce the training time.
For the other GNNs, optimization should focus on reducing the costs of the collect and the aggregation steps.
Additionally, improving the efficiency of the collect step can benefit all GNNs.

@@

\paragraph{GGNN}
The top basic operator was \texttt{mm} used in the vertex calculation.
Due to the high time complexity in $\gamma$, the proportion of the \texttt{mm} operator was much higher than the other operators.
The \texttt{thnn\_fused\_gru\_cell} operator that was used in the backward phase of $\gamma$ was also noticeable.
The other three operators were used in the edge calculation.

@@

\paragraph{Summary}
The GNN training is suitable for GPU.
The \textbf{edge calculation is the main performance bottleneck}, except for training GNNs with high vertex calculation complexity on low-average-degree graphs.
The performance bottleneck in the edge calculation depends on the time complexity of the message function $\phi$.
\begin{itemize}
    \item If the time complexity of $\phi$ is \textbf{high}, the \textbf{efficiency of $\phi$} limits the performance. Reducing its computation cost (via optimizing its implementation or modifying the algorithm) can significantly reduce the training time.
    \item If the time complexity of $\phi$ is \textbf{low}, the \textbf{collect step} and the \textbf{aggregation step} limit the performance. The collect step involves lots of data movement. The aggregation step suffers from data synchronization and non-regular data access. Optimizing their implementations can significantly reduce the training time.
\end{itemize}

@@

\figurename~\ref{fig:exp_memory_usage_stage_amp} shows the peak memory usage of each phase in the GNN training on the \texttt{amp} dataset.
The trend was similar to other datasets.
\emph{The GNN training achieved its peak memory usage in the forward and the backward phases}.
The forward phase generated a large number of intermediate results.
Some key intermediate results were cached, increasing memory usage.
The cached results were used in the gradient calculation in the backward phase.
\figurename~\ref{fig:ggnn_vertex_func_computation_graph} shows the computation graph of the vertex update function $\gamma$ of GGNN.
The computation graph had a large number of operators, each operator generating an intermediate matrix.
Some key intermediate matrices were cached.
The cached matrices were the main source of memory usage in the loss phase.
By the end of the backward phase, the cached results were released.
Since the evaluation phase did not need to calculate the gradients, it did not cache intermediate results.
Its memory usage declined sharply.

@@

\figurename~\ref{fig:exp_memory_expansion_ratio} also indicates that the same GNN had different MERs for different datasets.
Two characteristics of the dataset affect the MER: the dimension of the input feature vector and the average degree.

@@

The average degree of the graph also affects MER by influencing the relative scale of the intermediate results from the vertex/edge calculation.
Fixing the number of vertices $|\mathcal{V}|$, we used the R-MAT generator to generate random graphs with different average degrees.
\figurename~\ref{fig:exp_memory_expansion_ratio_input_graph_number_of_edges} shows how the memory usage changes according to the average degrees.
As the average degree $\bar{d}$ increased, the number of edges $|\mathcal{E}|$ increased and the peak memory usage increased \emph{linearly} with $\bar{d}$.
The edge calculation gradually dominated the memory usage and \emph{the MER converged to a stable value}.
The stable value was determined by the complexity of the edge calculation.
Except for GGNN, the MER of the other GNNs increased as $\bar{d}$ increased.
As GGNN had high vertex calculation complexity, the MER related to the vertex calculation was much larger than the edge calculation.
When the edge calculation dominated the memory usage, its MER became smaller.

@@

We also fixed the number of edges $|\mathcal{E}|$ in the graph and generated random graphs with different $|\mathcal{V}|$.
\figurename~\ref{fig:exp_memory_expansion_ratio_input_graph_number_of_vertices_fixed_edge} shows how the memory usage changed according to $|\mathcal{V}|$.
All GNNs were much more insensitive to the changes in $|\mathcal{V}|$ compared to $|\mathcal{E}|$.
Except for GGNN, the MER of the other GNNs declined as $|\mathcal{V}|$ increased because the scale of the dataset increased more quickly than the scales of the intermediate results.
As GGNN had high vertex calculation complexity, the scales of the intermediate results were much more sensitive to $|\mathcal{V}|$.
It indicates that \emph{the intermediate results of the edge calculation dominated the memory usage during the GNN training}.

@@

In the implementation in PyG, the GNN model and the dataset resided on the GPU side.
To process each epoch, PyG sampled the original dataset in the main memory and generated several batches.
Each batch was a small subgraph of the dataset.
To train on each batch, PyG sent the sampled subgraph to GPU, calculated the gradients in the subgraph, and updated the model parameters directly on GPU.
With the sampling techniques, the model parameters were updated by a stochastic gradient descent optimizer.
It conducted the evaluation phase every several epochs (either on the CPU side or the GPU side) to determine whether to stop the training.
The experiments focused on the training phase of each batch.

@@

\figurename~\ref{fig:exp_sampling_minibatch_graph_info} shows how the size of the sampled subgraph changed with the batch size.
For the neighbor sampler, the relative batch size is the proportion of the sampled vertices of the last GNN layer in all vertices of the graph.
For the cluster sampler, the relative batch size is the proportion of the sampled partitions to all partitions of the graph.
The neighbor sampler was very sensitive to the batch size.
As the batch size increased, the size of the sampled subgraph increased quickly and then stabilized.
The cluster sampler was much less sensitive compared to the neighbor sampler.
The number of vertices and the average degree of the sampled subgraphs increased linearly with the batch size.

@@

To find out the performance bottleneck with the sampling techniques, we decomposed the training time per batch into three phases: \emph{sampling} on CPU, \emph{transfering} sampled subgraphs from CPU to GPU and \emph{training} with the subgraphs on GPU.
\figurename~\ref{fig:exp_sampling_batch_train_time} shows the time breakdown of the four GNNs under different relative batch sizes.
For the neighbor sampler, the sampling technique reduced the training time per batch only when the batch size was very small.
When the batch became bigger, the sampling and the data transferring phases introduced noticeable overheads, even making the training time exceed the full-batch training.
For the clustering sampler, the sampled subgraph was smaller than the neighbor sampler under the same relative batch size.
The reduction in the training time was more obvious than the neighbor sampler.
However, the overheads increased quickly as the relative batch size increased.
The training time under the 25\% relative batch size already exceeded the time of full-batch training.
The experimental results indicate that the implementation of the sampling techniques in PyG is inefficient.
When the batch size was slightly big, more than 50\% of the time had been spent on sampling and data transferring.
\emph{The sampling techniques are only efficient under small batch sizes.}

@@

The disadvantage of the sampling technique is wasting GPU resources.
As the sampling techniques are only effective under small batch sizes, the sampled subgraphs will be very small in those cases.
They cannot make full use of the computing power of a GPU.
To simulate the cases of training GNNs with small sampled subgraphs, we generated random graphs with few vertices and measured the training time per epoch in \figurename~\ref{fig:exp_small_graph_train_time}.
As the number of vertices increased, the training time was almost unchanged except for GaAN.
The training time of GaAN increased only with $|\mathcal{V}| \geq 4000$.

\paragraph{Summary}
With small batch size, the sampling techniques could significantly reduce the training time per batch and the peak memory usage during the training.
However, small batch sizes could not make full use of the computing power of a GPU.
The sampled subgraphs had lower average degrees than the whole graph.
However, the current implementation of the sampling techniques was every inefficient under big batch sizes.
The time spent on the sampling phase and the data transferring phase could even exceed the training phase.

### 2020-07-23 Manuscript/sec5-insights.tex
Update section 1.
update by www.grammarly.com

\section{Insights}

@@

    \item \emph{Performance optimizations should focus on improving the efficiency of the edge calculation.}
          The edge calculation is the most time-consuming phase in most GNNs.
          \begin{itemize}
              \item If the complexity of the message function $\phi$ is high, the implementation of $\phi$ is critical to performance.
                    Improving its efficiency can significantly reduce the training time.
                    For example, the attention mechanism in GNNs (like GAT and GaAN) requires an extra sub-layer to calculate the attention weight of each edge.
                    Implementing it with the specially optimized basic operators on GPU is a potential optimization.
              \item If the complexity of $\phi$ is low, the efficiency of the collect step and the aggregation step becomes critical.
                    The existing GNN libraries \cite{DGL, PyG, ma2019_neugraph} already introduce the \emph{fused} operator to improve their efficiency.
                    When the message function $\phi$ is an assignment or a scalar multiplication of the hidden feature vector of the source vertex, the libraries replace the collect, message and aggregate steps with a single fused operator.
                    The fused operator calculates the aggregated vectors directly from the input hidden feature vectors, minimizing the memory footprints and overlapping the memory accessing with calculation.
                    By this way, it significantly reduces the training time of GNNs with low edge calculation complexity (like GCN) \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph}.
                    However, the applicable condition of the fused operator is very restrict.
                    It does not work for $\phi$ with more complex operations like matrix multiplication.
                    A potial optimization is proposing an implementation of the edge calculation that generates the aggregated vectors directly from the input hidden feature vectors on the fly, without materializing the parameter vectors and the message vectors.
          \end{itemize}
    \item \emph{The high memory usage caused by the intermediate results of the edge calculation limits the data scalability of the GNN training.}
          The memory expansion ratios of the typical GNNs are very high, making GPU unable to handle big graphs.
          To train GNNs with big datasets, one solution is to distribute the dataset among severl GPUs and frequently swap parts of the dataset between GPUs and the main memory \cite{ma2019_neugraph}.
          Another possible solution \cite{chen2016_training_deep} comes from the DNN training.
          It only checkpoints key intermediate results during the forward propagation and recalculates the missing results on demand during the backproporgation.
          Implementing the checkpoint mechanism in the GNN training is another potential optimization.

    \item \emph{Sampling techniques can significantly reduce the training time and the memory usage, but its implementation is still inefficient}.
          The sampling techniques are effective under small batch sizes.
          Its current implementation brings considerable overheads when the batch size becomes large.
          Improving the efficiency of the sampling is a potentil optimization.
          The subgraphs sampled with small batch sizes are small.
          They cannot make fully use of the computing power of a GPU.
          How to improve the GPU utilization under small batch sizes is another problem to solve.
          One possible solution is to train multiple batches asynchronously on the same GPU and use the asynchronous stochastic gradient descent to speed up the converge.

>>>>

\section{Insights}
\label{sec:insights}

@@

    \item \emph{Performance optimizations should focus on improving the efficiency of the edge calculation.}
          The edge calculation is the most time-consuming phase in most GNNs.
          \begin{itemize}
              \item If the complexity of the message function $\phi$ is high, the implementation of $\phi$ is critical to performance.
                    Improving its efficiency can significantly reduce training time.
                    For example, the attention mechanism in GNNs (like GAT and GaAN) requires an extra sub-layer to calculate the attention weight of each edge.
                    Implementing it with the specially optimized basic operators on GPU is a potential optimization.
              \item If the complexity of $\phi$ is low, the efficiency of the collect step and the aggregation step becomes critical.
                    The existing GNN libraries \cite{DGL, PyG, ma2019_neugraph} already introduce the \emph{fused} operator to improve their efficiency.
                    When the message function $\phi$ is an assignment or a scalar multiplication of the hidden feature vector of the source vertex, the libraries replace the collect, message, and aggregate steps with a single fused operator.
                    The fused operator calculates the aggregated vectors directly from the input hidden feature vectors, minimizing the memory footprints and overlapping the memory accessing with calculation.
                    In this way, it significantly reduces the training time of GNNs with low edge calculation complexity (like GCN) \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph}.
                    However, the applicable condition of the fused operator is very restricted.
                    It does not work for $\phi$ with more complex operations like matrix multiplication.
                    A potential optimization is proposing an implementation of the edge calculation that generates the aggregated vectors directly from the input hidden feature vectors on the fly, without materializing the parameter vectors and the message vectors.
          \end{itemize}
    \item \emph{The high memory usage caused by the intermediate results of the edge calculation limits the data scalability of the GNN training.}
          The memory expansion ratios of the typical GNNs are very high, making GPU unable to handle big graphs.
          To train GNNs with big datasets, one solution is to distribute the dataset among several GPUs and frequently swap parts of the dataset between GPUs and the main memory \cite{ma2019_neugraph}.
          Another possible solution \cite{chen2016_training_deep} comes from DNN training.
          It only checkpoints key intermediate results during the forward propagation and recalculates the missing results on demand during the backpropagation.
          Implementing the checkpoint mechanism in GNN training is another potential optimization.

    \item \emph{Sampling techniques can significantly reduce the training time and memory usage, but its implementation is still inefficient}.
          The sampling techniques are effective under small batch sizes.
          Its current implementation brings considerable overheads when the batch size becomes large.
          Improving the efficiency of the sampling is a potential optimization.
          The subgraphs sampled with small batch sizes are small.
          They cannot make full use of the computing power of a GPU.
          How to improve the GPU utilization under small batch sizes is another problem to solve.
          One possible solution is to train multiple batches asynchronously on the same GPU and use the asynchronous stochastic gradient descent to speed up the converge.
